# Set options

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
```

# Load IfH utility library

```{r}
remotes::install_github("tumbleowlee/ifhutil")
library(ifhutil)
ifh.init(c('numbers'), quiet = TRUE)
```

# Set path to `out.emapper.annotations` file from eggNOG

```{r}
annotated.file <- '../private/out.emapper.annotations'
```

# Generate mappings from eggNOG output

```{r}
# Read eggNOG-mapper annotation table
ifh.info('Read all inputs...')
annotated.df <- ifh.table.import(annotated.file, header = FALSE, sep = '\t', comment.char = '#', strip.white = TRUE, quote = '')
colnames(annotated.df) <- ifh.file.get.line(path = annotated.file, prefix = '#', comment.prefix = '##', sep = '\t')

# Rename 'query' to 'GENEID' for GSE
colnames(annotated.df)[colnames(annotated.df) == 'query'] <- 'GENEID'
colnames(annotated.df)[colnames(annotated.df) == 'Preferred_name'] <- 'PreferredName'
ifh.success('Annotated inputs loaded.')

# Expand the table based on 'GOs' and export to CSV
geneid.to.go <- ifh.table.expand(annotated.df, by = 'GOs', split = ',', limit.to = c('GENEID', 'PreferredName'), na = '-')
ifh.table.export(geneid.to.go, 'geneid-to-go.csv', row.names = FALSE)

# Expand the table based on 'KEGG_ko' and export to CSV
geneid.to.kegg <- ifh.table.expand(annotated.df, by = 'KEGG_ko', split = ',', limit.to = c('GENEID', 'PreferredName'), remove.prefix = 'ko:', na = '-')
ifh.table.export(geneid.to.kegg, 'geneid-to-kegg.csv', row.names = FALSE)
```

# Perform Gene-Set-Enrichment

This section describes all necessary steps to perform GSE for GO and KEGG.

## Create configuration

```{r}
# Setup default configuration
config <- data.frame(
    # Prevent any output
    quiet = TRUE,
    # Organism annotation database
    organism.annotation = "org.EcK12.eg.db",
    # Input file
    deseq.file = "../private/L12VLB12_MGaccession.csv",
    # Keytype: source of annotation
    key.type = "GO",
    # ont: "BP", "MF", "CC", "ALL"
    ont = "CC",
    # Number of permutations
    num.perm = 10000,
    # Minimum number of genes in set
    min.gs.size = 3,
    # Maximum number of genes in set
    max.gs.size = 800,
    # pvalue cutoff
    pvalue.cutoff = 0.05,
    # padjust method: "holm", "hochberg", "hommel", "bonferroni",
    #                 "BH", "BY", "fdr", "none"
    padjust.method = "none",
    # KEGG organism code, see https://www.genome.jp/kegg/catalog/org_list.html
    kegg.code = "ecok",
    # KEGG key type
    kegg.key.type = "kegg",
    # GO Map
    go.map = "geneid-to-go.csv",
    # KEGG Map
    kegg.map = "geneid-to-kegg.csv",
    # Count of categories to plot
    show.categories = 10000,
    # Pathway ID
    pathway.id = "ecok00010",
    pathway.gene.idtype = "KEGG"
)
```

## Install annotation database

```{r}
# Install/Load organism annotations
ifh.info("Load organism annotation...")
ifh.install(pkg = config$organism.annotation, quiet = config$quiet)
ifh.success("Organism annotation loaded.")
```

### List available keytypes

```{r}
ifh.info("Possible keytypes for", config$organism.annotation, ":")
ifh.info(keytypes(get(config$organism.annotation)))
```


## Load DESEQ CSV file

```{r}
# Read DESeq2 data
ifh.step("Load CSV data from", config$deseq.file, "...")
deseq2.df = ifh.table.import(config$deseq.file, header = TRUE, sep = ",")
ifh.success("Data loaded.")

ifh.step("Prepare gene list...")
orig.gene.list <- deseq2.df$log2FoldChange
names(orig.gene.list) <- deseq2.df$X
gene.list <- na.omit(orig.gene.list)
gene.list <- sort(gene.list, decreasing = TRUE)
ifh.success("Gene list prepared.")
```

## Execute GO GSE

### Load required libraries
```{r}
# Initialize and install dependencies
ifh.info("Check and install missing dependencies...")
ifh.init(c("clusterProfiler", "pathview", "enrichplot", "gridExtra",
           "DOSE", "ggridges", "ggplot2", "tools", "annotate", "GO.db", "HDO.db"), quiet = config$quiet)
```

### Perform GO GSE

```{r}
# Load cache file or import go-geneid mapping frame
go.geneid.df <- NA
if (config$go.map!= "") {
    ifh.info("Import GO map from", config$go.map)
    go.geneid.df <- ifh.table.import(config$go.map, header = TRUE, sep = "\t", comment.char = "#")
    ifh.success("GO map imported.")
}

# Run GO GSE with organism database or custom GO-GENEID mapping
ifh.step("Calculate GO gene set enrichment...")
ifh.quiet(config$quiet)
go.gse = NA
go.gene.list = NA

#if (config$go.map != "") {
    go.supported.ids <- select(get(config$organism.annotation), keys(get(config$organism.annotation), "GO"), "ENTREZID", "GO")[,'GO']
    go.ids.in.ont <- filterGOByOntology(go.supported.ids, ontology = c(config$ont))
    go.supported.ids <- go.supported.ids[go.ids.in.ont == TRUE]
    go.term.to.gene <- go.geneid.df[go.geneid.df$GO %in% go.supported.ids,]
    go.term.to.name <- go2term(go.geneid.df[,1])
    names(go.term.to.name) <- c('GO', 'NAME')

    go.gene.list <- gene.list

    go.gse = suppressMessages({
        GSEA(geneList = go.gene.list,
             exponent = 1,
             minGSSize = config$min.gs.size,
             maxGSSize = config$max.gs.size,
             pvalueCutoff = config$pvalue.cutoff,
             pAdjustMethod = config$padjust.method,
             verbose = FALSE,
             eps = 0,
             TERM2GENE = go.term.to.gene,
             TERM2NAME = go.term.to.name,
             by = "fgsea"
             )
    })

    convert_go_to_pref_name <- function(val, conversion_table) {
        values <- conversion_table$PreferredName[conversion_table$GENEID %in% val]
        values <- unique(values)
        return(paste(values, collapse = ',', sep = ','))
    }

    #mapping
    convert_list_go_to_pref_name <- function(list, conversion_table) {
        vec <- unlist(strsplit(list, "/"))
        gene_names <- sapply(vec, convert_go_to_pref_name, conversion_table = conversion_table)
        return(paste(gene_names, collapse = "/"))
    }

    #extract the dataframe from enrichKEGG object
    temp_df = go.gse@result

    #replace the IDs with sapply
    temp_df$core_enrichment = sapply(temp_df$core_enrichment, convert_list_go_to_pref_name, conversion_table = go.geneid.df)

    #modify the result table
    go.gse@result = temp_df

    for (key in names(go.gse@geneSets)) {
        go.gse@geneSets[key] <- sapply(go.gse@geneSets[key], convert_go_to_pref_name, conversion_table = go.geneid.df)
    }

    names(go.gse@geneList) <- sapply(names(go.gse@geneList), convert_go_to_pref_name, conversion_table = go.geneid.df)
    names(go.gene.list) <- sapply(names(go.gene.list), convert_go_to_pref_name, conversion_table = go.geneid.df)
#} else {
#     go.gene.list <- gene.list
#     go.gse = suppressMessages({
#         gseGO(geneList = go.gene.list,
#               ont = config$ont,
#               keyType = config$key.type,
#               #nPerm = 10000,
#               minGSSize = config$min.gs.size,
#               maxGSSize = config$max.gs.size,
#               pvalueCutoff = config$pvalue.cutoff,
#               verbose = FALSE,
#               OrgDb = config$organism.annotation,
#               pAdjustMethod = config$padjust.method)})
# }
ifh.quiet(FALSE)
ifh.step("Calculations for GO GSE finished.")
```

#### Plot GO GSE results

```{r}
ifh.step("Create GO dotplot...")
pdf("go.dotplot.pdf",width=12,height=8)
p <- dotplot(go.gse, showCategory = config$show.categories, split = ".sign", font.size=8) + facet_grid(.~.sign)
p
dev.off()
ifh.success("Dotplot created.")

ifh.step("Create GO cnetplot...")
pdf("go.cnetplot.pdf",width=24,height=16)
cnetplot(go.gse, categorySize="pvalue", color.params = list(foldChange=go.gene.list), showCategory = config$show.categories, colorEdge = TRUE, cex.params = list(category_label = 0.7, gene_label = 0.5))
dev.off()
ifh.success("Cnetplot created.")

ifh.step("Create GO heatplot...")
pdf("go.heatplot.pdf",width=70,height=18)
heatplot(go.gse, foldChange=go.gene.list, showCategory = config$show.categories)
dev.off()
ifh.success("Heatplot created.")
```

### Perform KEGG GSE

```{r}
# Load cache file or import KEGG-geneid mapping frame
kegg.geneid.df <- NA
if (config$kegg.map != "") {
    ifh.info("Import KEGG map from", config$kegg.map)
    kegg.geneid.df <- ifh.table.import(config$kegg.map, header = TRUE, sep = "\t", comment.char = "#")
    ifh.success("KEGG map imported.")
}

ifh.step("Prepare KEGG gene list...")
ifh.quiet(config$quiet)

# Run KEGG GSE with organism database or custom KEGG_ko-GENEID mapping
if (config$kegg.map != "") {
    kegg.df.unique.ids = kegg.geneid.df[!duplicated(kegg.geneid.df[c("GENEID")]),]
    deseq2.df = deseq2.df[deseq2.df$X %in% kegg.df.unique.ids[["GENEID"]],]
    kegg.df.unique.ids = kegg.df.unique.ids[kegg.df.unique.ids$GENEID %in% deseq2.df$X,]
    deseq2.df$Y = kegg.df.unique.ids$KEGG_ko

    kegg.gene.list <- deseq2.df$log2FoldChange
    names(kegg.gene.list) <- deseq2.df$Y
    kegg.gene.list <- na.omit(kegg.gene.list)
    kegg.gene.list = sort(kegg.gene.list, decreasing = TRUE)
    ifh.success("KEGG gene list prepared.")

    ifh.step("Calculate KEGG gene set enrichment...")
    kegg.gse <- gseKEGG(geneList = kegg.gene.list,
                        organism = "ko",
                        #nPerm = 10000,
                        minGSSize = config$min.gs.size,
                        maxGSSize = config$max.gs.size,
                        pvalueCutoff = config$pvalue.cutoff,
                        pAdjustMethod = config$padjust.method,
                        keyType = config$kegg.key.type,
                        eps = 0)

    convert_kegg_to_pref_name <- function(val, conversion_table) {
        values <- conversion_table$PreferredName[conversion_table$KEGG_ko %in% val]
        values <- unique(values)
        return(paste(values, collapse = ',', sep = ','))
    }

    #mapping
    convert_list_kegg_to_pref_name <- function(list, conversion_table) {
        vec <- unlist(strsplit(list, "/"))
        gene_names <- sapply(vec, convert_kegg_to_pref_name, conversion_table = conversion_table)
        return(paste(gene_names, collapse = "/"))
    }

    #extract the dataframe from enrichKEGG object
    temp_df = kegg.gse@result

    #replace the IDs with sapply
    temp_df$core_enrichment = sapply(temp_df$core_enrichment, convert_list_kegg_to_pref_name, conversion_table = kegg.geneid.df)

    #modify the result table
    kegg.gse@result = temp_df

    for (key in names(kegg.gse@geneSets)) {
        kegg.gse@geneSets[key] <- sapply(kegg.gse@geneSets[key], convert_kegg_to_pref_name, conversion_table = kegg.geneid.df)
    }

    names(kegg.gse@geneList) <- sapply(names(kegg.gse@geneList), convert_kegg_to_pref_name, conversion_table = kegg.geneid.df)
} else {
    ids <- bitr(names(orig.gene.list), fromType = config$key.type, toType = "ENTREZID", OrgDb = config$organism.annotation)

    kegg.df.unique.ids = ids[!duplicated(ids[c(config$key.type)]),]
    deseq2.df = deseq2.df[deseq2.df$X %in% kegg.df.unique.ids[[config$key.type]],]
    deseq2.df$Y = kegg.df.unique.ids$ENTREZID

    kegg.gene.list <- deseq2.df$log2FoldChange
    names(kegg.gene.list) <- deseq2.df$Y
    kegg.gene.list <- na.omit(kegg.gene.list)
    kegg.gene.list = sort(kegg.gene.list, decreasing = TRUE)
    ifh.success("KEGG gene list prepared.")

    ifh.step("Calculate KEGG gene set enrichment...")
    kegg.gse <- gseKEGG(geneList = kegg.gene.list,
                        organism = config$kegg.code,
                        #nPerm = 10000,
                        minGSSize = config$min.gs.size,
                        maxGSSize = config$max.gs.size,
                        pvalueCutoff = config$pvalue.cutoff,
                        pAdjustMethod = config$padjust.method,
                        keyType = config$kegg.key.type)
}
ifh.quiet(FALSE)
ifh.step("Calculations for KEGG GSE finished.")
```

#### Plot KEGG GSE results

```{r}
ifh.step("Create KEGG dotplot...")
pdf("kegg.dotplot.pdf",width=12,height=8)
dotplot(kegg.gse, showCategory = config$show.categories, title = "Enriched Pathways", split = ".sign", font.size=8) + facet_grid(.~.sign)
dev.off()
ifh.success("Dotplot created.")

ifh.step("Create KEGG cnetplot...")
pdf("kegg.cnetplot.pdf",width=36,height=12)
cnetplot(kegg.gse, categorySize="pvalue", color.params = list(foldChange = kegg.gse@geneList, edge = TRUE), showCategory = config$show.categories, cex.params = list(category_label = 0.7, gene_label = 0.5))
dev.off()
ifh.success("Cnetplot created.")

ifh.step("Create KEGG heatplot...")
pdf("kegg.heatplot.pdf",width=50,height=16)
heatplot(kegg.gse, foldChange=kegg.gse@geneList, showCategory = config$show.categories)
dev.off()
ifh.success("Heatplot created.")

ifh.step("Create KEGG pathview...")
pathview(cpd.data=kegg.gene.list, pathway.id = config$pathway.id, species = config$kegg.code, gene.idtype = config$pathway.gene.idtype)
ifh.success("Pathview created.")
```
